---
title: "Sales Data Visualizations (Base R Only)"
author: "Jacob Siau"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
---

```{r setup, include=FALSE}
# Base R only: no external packages required
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Load Data (Base R)

```{r}
# Adjust the path if your CSV is elsewhere
csv_path <- "sales_data_sample.csv"

sales <- read.csv(csv_path, stringsAsFactors = FALSE)

# Ensure expected columns exist
stopifnot(all(c("OrderID","OrderDate","Region","Product","Quantity","UnitPrice","TotalSales") %in% names(sales)))

# Parse dates
sales$OrderDate <- as.Date(sales$OrderDate)

# Convert Region/Product to factors for plotting
sales$Region  <- factor(sales$Region)
sales$Product <- factor(sales$Product)

str(sales)
summary(sales$OrderDate)
```

# 1) Bar graph: Sum of Sales by Product (Base R)

```{r fig.width=8, fig.height=5}
# Aggregate total sales by product
sales_by_product <- aggregate(TotalSales ~ Product, data = sales, FUN = function(x) sum(x, na.rm = TRUE))

# Order descending for nicer plot
ord <- order(sales_by_product$TotalSales, decreasing = TRUE)
sales_by_product <- sales_by_product[ord, ]

par(mar = c(5, 9, 4, 2) + 0.1)
barplot(height = sales_by_product$TotalSales,
        names.arg = sales_by_product$Product,
        horiz = TRUE, las = 1,
        main = "Total Sales by Product",
        xlab = "Sum of Sales")
```

# 2) Pie chart: Sum of Sales by Region (Base R)

```{r fig.width=6, fig.height=6}
sales_by_region <- tapply(sales$TotalSales, sales$Region, function(x) sum(x, na.rm = TRUE))

# Make labels with percentages
region_prop <- sales_by_region / sum(sales_by_region)
labels <- paste0(names(sales_by_region), " (", round(region_prop * 100), "%)")

pie(sales_by_region,
    labels = labels,
    main = "Share of Total Sales by Region")
```

# 3) Box Plots: Order-level Sales and Quantities by Region (Base R)

```{r fig.width=10, fig.height=4}
par(mfrow = c(1, 2))

boxplot(TotalSales ~ Region, data = sales,
        main = "Order Total Sales by Region",
        xlab = "Region", ylab = "Total Sales", outline = TRUE)

boxplot(Quantity ~ Region, data = sales,
        main = "Order Quantities by Region",
        xlab = "Region", ylab = "Quantity", outline = TRUE)

par(mfrow = c(1, 1))
```

# 4) Histogram/Bar: Count of Orders by Product (Base R)

```{r fig.width=8, fig.height=5}
counts_by_product <- table(sales$Product)

par(mar = c(5, 9, 4, 2) + 0.1)
barplot(counts_by_product[order(counts_by_product, decreasing = TRUE)],
        horiz = TRUE, las = 1,
        main = "Number of Orders by Product",
        xlab = "Order Count")
```

# 5) Radial Plot: Total Sales by Month (All Products) — Base R, No Packages

```{r fig.width=7, fig.height=7}
# Compute monthly totals (calendar months across all years)
# Use month abbreviations and keep them in Jan..Dec order
month_levels <- month.abb  # built-in constant
sales$MonthAbb <- factor(format(sales$OrderDate, "%b"), levels = month_levels)

monthly_totals <- tapply(sales$TotalSales, sales$MonthAbb, function(x) sum(x, na.rm = TRUE))
monthly_totals[is.na(monthly_totals)] <- 0

# Normalize radii to [0, 1] for plotting
r <- as.numeric(monthly_totals)
if (all(r == 0)) {
  r_scaled <- rep(0, length(r))
} else {
  r_scaled <- r / max(r)
}

# Angles for 12 months, start at 90 degrees (up) and go clockwise
theta <- seq(from = pi/2, to = pi/2 - 2*pi + (2*pi/12), length.out = 12)
# Or equivalently: theta <- (pi/2) - 2*pi*(0:11)/12

# Plot setup
par(pty = "s")  # square plot
plot(NA, NA, xlim = c(-1.1, 1.1), ylim = c(-1.1, 1.1),
     xlab = "", ylab = "", axes = FALSE,
     main = "Total Sales by Month (Radial)")

# Draw circular grid
symbols(0, 0, circles = c(1, 0.75, 0.5, 0.25), inches = FALSE, add = TRUE, fg = "grey80")

# Draw spokes and points
for (i in 1:12) {
  x <- r_scaled[i] * cos(theta[i])
  y <- r_scaled[i] * sin(theta[i])
  # spoke
  segments(0, 0, x, y, col = "grey40")
  # point at the end of spoke
  points(x, y, pch = 16, cex = 1.2)
}

# Connect the months to form a radial polygon
xp <- r_scaled * cos(theta)
yp <- r_scaled * sin(theta)
polygon(xp, yp, border = "grey30", col = rgb(0.3, 0.3, 0.8, 0.2))

# Month labels slightly beyond unit circle
label_radius <- 1.05
for (i in 1:12) {
  text(label_radius * cos(theta[i]), label_radius * sin(theta[i]),
       labels = month_levels[i], cex = 0.9)
}

# Add a simple legend for scale (max value)
legend("bottomleft",
       legend = paste0("Max monthly total: ", format(max(r), big.mark = ",")),
       bty = "n")
```

## Notes
- This document relies only on base R functions (`read.csv`, `aggregate`, `tapply`, `barplot`, `boxplot`, `pie`, `segments`, `polygon`, etc.).
- Ensure your `OrderDate` column is in a format recognized by `as.Date`. If needed, specify a format like: `as.Date(sales$OrderDate, format = "%Y-%m-%d")`.
- If you want a true polar bar chart, you typically need an additional package. The above radial plot approximates a polar/radar style using base graphics.
